---
title: "Modelling - El Paso"
author: "Sisun Cheng, Kristin Chang, Jenna Epstein"
date: "2022/3"
output:
  html_document:
    toc: yes
    toc_float: TRUE
    code_folding: hide
editor_options:
  chunk_output_type: inline
---
# 0. Set Up

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, results = FALSE, message = FALSE)
```

```{r libraries, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(tidycensus)
library(gganimate)
library(viridis)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(RSocrata)
library(caret)
library(purrr)
library(FNN)
library(stargazer)
library(dplyr)
library(spatstat)
library(raster)
library(spdep)
library(grid)
library(mapview)
library(stringr)
library(ggcorrplot)
library(scales)
library(colorspace)
library(rgdal)          
library(RColorBrewer) 
library(rasterVis)    
library(sp)
library(ggpubr)
library(leaflet)
library(gganimate)
library(gifski)
library(transformr)
library(jtools)
library(stringr)

library(randomForest)
library(e1071)  # SVM
library(xgboost)
library(readr)
library(car)

palette_5 <- c("#0c1f3f", "#08519c", "#3bf0c0", "#e6a52f", "#e76420")
palette_5blues <-c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette_4 <-c("#08519c","#3bf0c0","#e6a52f","#e76420")
palette_2 <-c("#e6a52f","#08519c")
palette_3 <-c("#e6a52f","#08519c", "#e76420")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
```

```{r add plot and map themes for default visualizations, message=FALSE, warning=FALSE}
mapTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle=element_text(face="italic"),
    plot.caption=element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),axis.title = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.text.x = element_text(size = 14))
}

plotTheme <- function(base_size = 12) {
  theme(
    text = element_text( color = "black"),
    plot.title = element_text(size = 16,colour = "black"),
    plot.subtitle = element_text(face="italic"),
    plot.caption = element_text(hjust=0),
    axis.ticks = element_blank(),
    panel.background = element_blank(),
    panel.grid.major = element_line("grey80", size = 0.1),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", fill=NA, size=2),
    strip.background = element_rect(fill = "grey80", color = "white"),
    strip.text = element_text(size=12),
    axis.title = element_text(size=12),
    axis.text = element_text(size=10),
    plot.background = element_blank(),
    legend.background = element_blank(),
    legend.title = element_text(colour = "black", face = "italic"),
    legend.text = element_text(colour = "black", face = "italic"),
    strip.text.x = element_text(size = 14)
  )
}

```

Please download the **EDA.RData** file from [google drive](https://drive.google.com/file/d/1sr7AMW5YAfoxHK-9UVq9fOxVcafevpJp/view?usp=sharing) and put it under Data/, I've added the path to gitignore file.

```{r import rdata from EDA, message=FALSE, warning=FALSE}
# RData from EDA part, please make sure to run this properly
 load("Data/EDA.RData")

# OPTIONAL BELOW:
# modelling_0325.RData contains all the model in this Rmd file, if you do not wish to run through all the models below and just want to take a quick look at the model results, you can download the RData file here(https://drive.google.com/file/d/15yaeGVxl-Wnh78o9jxEzRJgCwMT7taa8/view?usp=sharing), save it in the corresponding root and load it. 
#load("Data/modelling_0325.RData")
```

# 1. Further EDA and Feature Engineering
In this part, we work on some further exploratory analysis and feature engineering since last markdown and presentation.
## OSM Amenities
We gather the amenity features from python codes.
```{r load amentiy data from outside source, message=FALSE, warning=FALSE}
amenity <- read.csv("Data/OSM_amenities/OSM_amenity.csv")

amenity_sf <- 
  amenity %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant") %>%
  st_transform('ESRI:102339')
```


```{r plot the amenities, message=FALSE, warning=FALSE}
ggplot() +
  geom_sf(data = El_Paso_city, color = "grey") +
  geom_sf(data = amenity_sf, aes(color = amenity), size=0.5, show.legend = "point") +
  labs(title = "Open Street Map Amenities",
       subtitle = "El Paso, TX")+
  mapTheme()

ggplot(amenity_sf, aes(y=amenity)) +
  geom_bar(width=0.5, color="black", fill = "#08519c") +
  labs(title = "Amenities by Type",
       x="Count",
       y="Type",
       subtitle = "El Paso, TX") + plotTheme()
```

```{r divide amenities to different categories, message=FALSE, warning=FALSE}
food_drink <- c('restaurant', 'fast_food', 'cafe', 'bar', 'ice_cream', 'pub')
entertainment <- c('arts_centre', 'cinema', 'theatre')
car_facility <- c('fuel', 'car_rental', 'car_wash', 'parking', 'parking_space')

am_food_drink <- amenity_sf %>%
  filter(amenity %in% food_drink)
am_entertainment <- amenity_sf %>%
  filter(amenity %in% entertainment)
am_car_facility <- amenity_sf %>%
  filter(amenity %in% car_facility)

```


```{r join the amenities to main df, message=FALSE, warning=FALSE}
# KNN function
st_c <- st_coordinates
EPCenterline_new6 <-
  EPCenterline_new5 %>%
  mutate(
      food_drink_nn3 = nn_function(na.omit(st_c(st_centroid(EPCenterline_new5))),na.omit(st_c(am_food_drink)), 3),
      entertainment_nn3 = nn_function(na.omit(st_c(st_centroid(EPCenterline_new5))),na.omit(st_c(am_entertainment)), 3),
      car_facility_nn3 = nn_function(na.omit(st_c(st_centroid(EPCenterline_new5))),na.omit(st_c(am_car_facility)), 3),
  )

```

```{r test correlation for PCI and knns, message=FALSE, warning=FALSE}
EPCenterline_new6 %>% 
  st_drop_geometry() %>%
  dplyr::select(PCI_2018, food_drink_nn3, entertainment_nn3, car_facility_nn3) %>%
  gather(Variable, Value, -PCI_2018) %>% 
  ggplot(aes(Value, PCI_2018)) +
  geom_point(size = 0.5, color = "grey") + 
  geom_smooth(method = "lm", se=F, colour = "#3FB0C0") +
     facet_wrap(~Variable, ncol = 3, scales = "free") +
     labs(title = "Nearest Neighbor Variables vs. PCI",
          subtitle = "El Paso, TX") + 
  stat_cor(aes(label = ..r.label..), label.x = 0) +
  plotTheme()
```

## Flood Zones

```{r bring in the 2020 preliminary flood zone data, message=FALSE, warning=FALSE}
prelim_floodzones_2020 <- st_read("Data/PrelimFloodZone2020/PrelimFloodZone2020.shp" )%>%
  st_transform('ESRI:102339')
```


```{r clip prelim flood zones to city boundaries, message=FALSE, warning=FALSE}
# clip flood zones to city bounds
EPcity_prelim_floodzones_2020 <- st_intersection(prelim_floodzones_2020, El_Paso_city)

EPcity_prelim_floodzones_2020 <- EPcity_prelim_floodzones_2020 %>% dplyr::filter(EPcity_prelim_floodzones_2020$STUDY_TYP == "SFHAs WITH HIGH FLOOD RISK")

```


```{r visualize flood zones, message=FALSE, warning=FALSE}
ggplot()+
  geom_sf(data=El_Paso_city, aes(), color="grey")+
  geom_sf(data = EPcity_prelim_floodzones_2020, aes(fill=STUDY_TYP), color="transparent", fill="#3FB0C0")+
  #scale_fill_viridis_d(direction=-1, option='G')+
  labs(title="Preliminary Flood Zone 2020: Area of High Flood Risk",
       fill= "",
       subtitle="El Paso, TX", caption="Source: City of El Paso; FEMA") + mapTheme()

```


```{r flood zones by high risk, message=FALSE, warning=FALSE}
EPCenterline_new6 <- EPCenterline_new6 %>% mutate(n_floodzone_int = lengths(st_intersects(EPCenterline_new6, EPcity_prelim_floodzones_2020)))

# adding a column for yes or no for if intersected a flood zone area at all or not
EPCenterline_new6 <-
  EPCenterline_new6 %>%
  mutate(floodzone_highrisk = ifelse(EPCenterline_new6$n_floodzone_int > 0, "Yes", "No"))
```

## Road Age Dummy

```{r Testing out a dummy variable for ROAD AGE}
# Create dummy variable
EPCenterline_new6$road_age_dummy <- ifelse(EPCenterline_new6$road_age >= 20, 1, 0)

# run a baby model
#OLS_reg_roadageonly <- lm(PCI_2018 ~ road_age_dummy, data = EPCenterline_new6)

#stargazer(OLS_reg_roadageonly, type = "text",title = "OLS Regression Results", align=TRUE, no.space=TRUE)
#summary(OLS_reg_roadageonly)
```

## Crash Updates
```{r udpated crash data according to year, message=FALSE, warning=FALSE}
#read in new data
crash16 <- st_read("Data/CRIS2016/CRIS2016.shp")
crash17<- st_read("Data/CRIS2017/CRIS2017.shp")
crash19<- st_read("Data/CRIS2019/CRIS2019.shp")
crash20<- st_read("Data/CRIS2020/CRIS2020.shp")
crash21<- st_read("Data/CRIS2021/CRIS2021.shp")

#combine past years
crash16_18 <- rbind(crash16, crash17, crash18)
crash19_21 <- rbind(crash19, crash20, crash21)

#replace 0s in lat long columns with NA so we can omit
crash16_18trim<-crash16_18[!(crash16_18$Latitude==0 | crash16_18$Longitude==0),]
crash19_21trim<-crash19_21[!(crash19_21$Latitude==0 | crash19_21$Longitude==0),]

#transforming to our crs
crash16_18sf <- crash16_18trim %>%
  na.omit() %>%
  st_as_sf(coords = c("Latitude", "Longitude"),
           crs = 'epsg:2277',
           agr = "constant") %>%
  st_transform('ESRI:102339')

crash19_21sf <- crash19_21trim %>%
  na.omit() %>%
  st_as_sf(coords = c("Latitude", "Longitude"),
           crs = 'epsg:2277',
           agr = "constant") %>%
  st_transform('ESRI:102339')
```

```{r join crash data to EPcenterline, message=FALSE, warning=FALSE}
#join crashes to EPCenterline using nearest feature 
crash_centerlines16_18 <-  st_join(crash16_18sf, EPCenterline_buffer, join = st_nearest_feature)
crash_centerlines19_21 <-  st_join(crash19_21sf, EPCenterline_buffer, join = st_nearest_feature)

#clean up to make it easier
crash_centerlines_clean16_18 <- crash_centerlines16_18 %>%
  dplyr::select(Crash_ID, index) %>% st_drop_geometry()

crash_centerlines_clean19_21 <- crash_centerlines19_21 %>%
  dplyr::select(Crash_ID, index) %>% st_drop_geometry()

#drop old columns
EPCenterline_new6 <- subset(EPCenterline_new6, select= -c(crash_count,crash_len))

# count crashes per street segment
crash_groupings16_18 <- crash_centerlines_clean16_18 %>%
  group_by(index) %>%
  summarize(crash_count16_18=n())

crash_groupings19_21 <- crash_centerlines_clean19_21 %>%
group_by(index) %>%
summarize(crash_count19_21=n())
```


```{r join crash data to EPcenterline for 2016 to 2018, message=FALSE, warning=FALSE}
# for 2016-2018...

# then join back to initial EPCenterline using index as the ID
EPCenterline_new6 <- merge(EPCenterline_new6, crash_groupings16_18, by = "index", all.x=TRUE)

# replace NAs in crash count column with 0
EPCenterline_new6$crash_count16_18[is.na(EPCenterline_new6$crash_count16_18)] <- 0

# calculate crashes per 100 meters
EPCenterline_new6 <-
  EPCenterline_new6 %>%
  mutate(crash_len16_18 = crash_count16_18*100/pave_length)

# convert to numeric
EPCenterline_new6$crash_len16_18 <- as.numeric(as.character(EPCenterline_new6$crash_len16_18))
```


```{r join crash data to EPcenterline for 2019 to 2021, message=FALSE, warning=FALSE}
# then again for 2019-2021...
# then join back to initial EPCenterline using index as the ID
EPCenterline_new6 <- merge(EPCenterline_new6, crash_groupings19_21, by = "index", all.x=TRUE)

# replace NAs in crash count column with 0
EPCenterline_new6$crash_count19_21[is.na(EPCenterline_new6$crash_count19_21)] <- 0

# calculate crashes per 100 meters
EPCenterline_new6 <-
  EPCenterline_new6 %>%
  mutate(crash_len19_21 = crash_count19_21*100/pave_length)

# convert to crash_len19_21
EPCenterline_new6$crash_len19_21 <- as.numeric(as.character(EPCenterline_new6$crash_len19_21))
```

## Roadbed Features

```{r join roadbed feats to centerlines, message=TRUE, warning=TRUE}
#rename target column and drop unneccessary columns
roadbed_base_PCI <- roadbed_base_PCI %>%
  mutate(rb_base = BASE_TYPE_)%>%
  dplyr::select(index, rb_base)%>%
  st_drop_geometry()

roadbed_surface_PCI <- roadbed_surface_PCI %>%
  mutate(rb_surface = SRFC_TYPE)%>%
  dplyr::select(index, rb_surface)%>%
  st_drop_geometry()

#get rid of duplicates - selecting first value of the group (duplicates have same value)
roadbed_base_PCI <- roadbed_base_PCI %>%
  group_by(index)%>%
  summarize(rb_base = first(rb_base))

roadbed_surface_PCI <- roadbed_surface_PCI %>%
  group_by(index)%>%
  summarize(rb_surface = first(rb_surface))

roadbed_surface_PCI$rb_surface <- str_to_title(roadbed_surface_PCI$rb_surface)

#join to EPCenterline_new6
EPCenterline_new6 <- left_join(EPCenterline_new6, roadbed_base_PCI, on="index")
EPCenterline_new6 <- left_join(EPCenterline_new6, roadbed_surface_PCI, on="index")

#change NA values to unknown category
EPCenterline_new6[,c("rb_base","rb_surface")][is.na(EPCenterline_new6[,c("rb_base","rb_surface")])] <- "Unknown"
```

## Potholes Update
Creating new columns for potholes counts and lengths for the two different year groupings

```{r update potholes feature, message=TRUE, warning=TRUE}
#drop old columns
EPCenterline_new6 <- subset(EPCenterline_new6, select= -c(potholes_count,potholes_len))

#clean up potholes grouping dfs
potholes_sf_2016to2018 <- subset(potholes_sf_2016to2018, select= c(YEAR, geometry, WORKORDERID))
potholes_sf_2019to2021 <- subset(potholes_sf_2019to2021, select= c(YEAR, geometry, WORKORDERID))

```

```{r join potholes to our df, message=TRUE, warning=TRUE}
#join potholes to EPCenterline using nearest feature 
potholes_centerlines_2016to2018 <-  st_join(potholes_sf_2016to2018, EPCenterline_buffer, join = st_nearest_feature)
potholes_centerlines_2019to2021 <-  st_join(potholes_sf_2019to2021, EPCenterline_buffer, join = st_nearest_feature)

#clean up to make it easier
potholes_centerlines_clean_2016to2018 <- potholes_centerlines_2016to2018 %>%
  dplyr::select(WORKORDERID, index) %>% st_drop_geometry()

potholes_centerlines_clean_2019to2021 <- potholes_centerlines_2019to2021 %>%
  dplyr::select(WORKORDERID, index) %>% st_drop_geometry()

# count potholes per street segment
potholes_groupings_2016to2018 <- potholes_centerlines_clean_2016to2018 %>%
  group_by(index) %>%
  summarize(potholes_count16_18=n())

potholes_groupings_2019to2021 <- potholes_centerlines_clean_2019to2021 %>%
group_by(index) %>%
summarize(potholes_count19_21=n())

```

```{r potholes 2016-2018, message=TRUE, warning=TRUE}
# for 2016-2018...

# then join back to initial EPCenterline using index as the ID
EPCenterline_new6 <- merge(EPCenterline_new6, potholes_groupings_2016to2018, by = "index", all.x=TRUE)

# replace NAs in potholes count column with 0
EPCenterline_new6$potholes_count16_18[is.na(EPCenterline_new6$potholes_count16_18)] <- 0

# calculate potholes per 100 meters
EPCenterline_new6 <-
  EPCenterline_new6 %>%
  mutate(potholes_len16_18 = potholes_count16_18*100/pave_length)

# convert to numeric
EPCenterline_new6$potholes_len16_18 <- as.numeric(as.character(EPCenterline_new6$potholes_len16_18))
```

```{r potholes 2019-2021, message=TRUE, warning=TRUE}
# for 2019-2021...

# then join back to initial EPCenterline using index as the ID
EPCenterline_new6 <- merge(EPCenterline_new6, potholes_groupings_2019to2021, by = "index", all.x=TRUE)

# replace NAs in potholes count column with 0
EPCenterline_new6$potholes_count19_21[is.na(EPCenterline_new6$potholes_count19_21)] <- 0

# calculate potholes per 100 meters
EPCenterline_new6 <-
  EPCenterline_new6 %>%
  mutate(potholes_len19_21 = potholes_count19_21*100/pave_length)

# convert to numeric
EPCenterline_new6$potholes_len19_21 <- as.numeric(as.character(EPCenterline_new6$potholes_len19_21))
```

```{r functional classification}
func_class <- st_read("Data/ElPaso_Func_Class/ElPaso_Func_Class.shp")%>%
  st_transform('ESRI:102339')

major_FCs <- c("Major Collector", "Principal Arterial - Other", "Interstate", "Principal Arterial - (Other Freeways and Expressways)")

major_arterials <- func_class%>%
  filter(FC_DESC == major_FCs)

#return index of nearest feat in major_arterials to each feature in EPCenterline
nearest_major_arterial <- st_nearest_feature(EPCenterline_new6, major_arterials) 

EPCenterline_new6 <- EPCenterline_new6 %>%
  mutate(dist_major_arterial = st_distance(EPCenterline_new6, major_arterials[nearest_major_arterial,], by_element=TRUE))
```


# 2. Modelling

## Splitting Data according to year
```{r EPCenterline data for 2016 to 2018 that includes PCI 2018, message=TRUE, warning=TRUE}
# 2016-2018
EPCenterline_2016to2018<- subset(EPCenterline_new6, select= -c(crash_count19_21, crash_len19_21, potholes_count19_21,potholes_len19_21))

```

```{r EPCenterline data for 2019 to 2021 that we will PREDICT ON, message=TRUE, warning=TRUE}
# 2019-2021
EPCenterline_2019to2021 <- subset(EPCenterline_new6, select= -c(crash_count16_18, crash_len16_18, potholes_count16_18,potholes_len16_18, PCI_2018))

```

## Variable Transformation

```{r distribution of predictors, message=TRUE, warning=TRUE}
EP_model <- EPCenterline_2016to2018 %>% na.omit()
hist(EP_model$crash_len16_18)
hist(EP_model$potholes_len16_18)

hist(EP_model$car_facility_nn3)
hist(EP_model$entertainment_nn3)
hist(EP_model$food_drink_nn3)

hist(EP_model$road_age)
hist(EP_model$VMT_pop)
hist(EP_model$dist_hydro)
hist(EP_model$dist_major_int)
```

```{r log transformation, message=TRUE, warning=TRUE}
EP_model$crash_len16_18 <- log(EP_model$crash_len16_18 + 2)
hist(EP_model$crash_len16_18)
EP_model$potholes_len16_18 <- log(EP_model$potholes_len16_18 + 2)
hist(EP_model$potholes_len16_18)

EP_model$dist_hydro <- log(as.numeric(EP_model$dist_hydro) + 2)
hist(EP_model$dist_hydro)

```

## Split to train and test set

```{r split the data to train and test set, message=TRUE, warning=TRUE}
set.seed(111)
# EP_model <- EP_model %>% na.omit()

EP_model$ind <- sample(2, nrow(EP_model), replace = TRUE, prob=c(0.8, 0.2))

EP_model_train <-
  EP_model %>% 
  subset(ind == 1) 

EP_model_test <-
  EP_model %>% 
  subset(ind == 2) 
```

## OLS Regression

```{r start with linear regression, message=TRUE, warning=TRUE}
OLS_reg1 <- 
  lm(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + #waze_count + 
       car_facility_nn3 + entertainment_nn3 + #food_drink_nn3 + 
       road_age + VMT_pop + dist_hydro + dist_major_int + 
       # med_hh_income + pct_transport_to_work + pctWhite + 
       CLASS + land_use_type + floodzone_highrisk,
     data = EP_model_train)

stargazer(OLS_reg1, type = "text",title = "OLS Regression Results", align=TRUE, no.space=TRUE)
summary(OLS_reg1)

EP_model_OLS <-
  EP_model_test %>%
  mutate(PCI.Predict = predict(OLS_reg1, EP_model_test),
         PCI.Error = PCI.Predict - PCI_2018,
         PCI.AbsError = abs(PCI.Predict - PCI_2018),
         PCI.APE = (abs(PCI.Predict - PCI_2018)) / PCI_2018) 

MAE <- mean(EP_model_OLS$PCI.AbsError, na.rm = T)
MAPE <- mean(EP_model_OLS$PCI.APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

```{r k-fold cross validation for OLS model, message=TRUE, warning=TRUE}
fitControl <- trainControl(method = "cv", number = 100)
set.seed(825)

OLS_reg1.cv <- 
  train(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + #waze_count + 
         car_facility_nn3 + entertainment_nn3 + #food_drink_nn3 + 
         road_age + VMT_pop + dist_hydro + dist_major_int + 
         # med_hh_income + pct_transport_to_work + pctWhite + 
         CLASS + land_use_type,
       data = EP_model_train,
        method = "lm", trControl = fitControl, na.action = na.pass)

OLS_reg1.cv
OLS_reg1.cv$resample[1:10,]
```

## Poisson Regression

```{r try poisson regression, message=TRUE, warning=TRUE}
poisson_reg1 <- glm(formula = 
                      PCI_2018 ~  crash_len16_18 + potholes_len16_18 + #waze_count + 
                       car_facility_nn3 + entertainment_nn3 + #food_drink_nn3 + 
                       road_age + VMT_pop + dist_hydro + dist_major_int + 
                       # med_hh_income + pct_transport_to_work + pctWhite + 
                       CLASS + land_use_type, 
                    data = EP_model_train, 
                    family = poisson)
stargazer(poisson_reg1, type = "text",title = "Poisson Regression Results", align=TRUE, no.space=TRUE)

EP_model_poisson <-
  EP_model_test %>%
  mutate(PCI.Predict = predict(poisson_reg1, EP_model_test),
         PCI.Error = PCI.Predict - PCI_2018,
         PCI.AbsError = abs(PCI.Predict - PCI_2018),
         PCI.APE = (abs(PCI.Predict - PCI_2018)) / PCI_2018) 

MAE <- mean(EP_model_poisson$PCI.AbsError, na.rm = T)
MAPE <- mean(EP_model_poisson$PCI.APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

## Random Forest

```{r set up rf model1, message=TRUE, warning=TRUE}
rf_model1 <- randomForest(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                           car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                           road_age + VMT_pop + dist_hydro + dist_major_int + 
                           CLASS + land_use_type + 
                           rb_base + rb_surface +
                           floodzone_highrisk,
                         data = EP_model_train, 
                         ntree=750,
                         mtry=15,
                         bootstrap=T)

rf_model1
importance(rf_model1)
```

```{r set up rf model2, message=TRUE, warning=TRUE}
rf_model2 <- randomForest(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                           car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                           road_age + VMT_pop + dist_hydro + dist_major_int + 
                           CLASS + land_use_type + 
                           rb_base + rb_surface +
                           floodzone_highrisk,
                         data = EP_model_train, 
                         ntree=750,
                         mtry=15,
                         bootstrap=F)

rf_model2
importance(rf_model2)
```

```{r prediction with rf model 1, message=TRUE, warning=TRUE}
EP_model_rf1 <- 
  EP_model_test %>%
  mutate(rf_predict = predict(rf_model1, EP_model_test),
         rf_error = rf_predict - PCI_2018,
         rf_absError = abs(rf_predict - PCI_2018),
         rf_APE = (abs(rf_predict - PCI_2018)) / PCI_2018) 


MAE <- mean(EP_model_rf1$rf_absError, na.rm = T)
MAPE <- mean(EP_model_rf1$rf_APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

```{r prediction with rf model 2, message=TRUE, warning=TRUE}
EP_model_rf2 <- 
  EP_model_test %>%
  mutate(rf_predict2 = predict(rf_model2, EP_model_test),
         rf_error2 = rf_predict2 - PCI_2018,
         rf_absError2 = abs(rf_predict2 - PCI_2018),
         rf_APE2 = (abs(rf_predict2 - PCI_2018)) / PCI_2018) 


MAE <- mean(EP_model_rf2$rf_absError2, na.rm = T)
MAPE <- mean(EP_model_rf2$rf_APE2, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

```{r grid search mtry for RF model, message=TRUE, warning=TRUE}
control <- trainControl(method="repeatedcv", number=10, repeats=3, search="grid")
set.seed(111)
tunegrid <- expand.grid(.mtry=c(1, 5, 10, 15))
rf_gridsearch <- train(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                           car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                           road_age + VMT_pop + dist_hydro + dist_major_int + 
                           CLASS + land_use_type + 
                           rb_base + rb_surface +
                           floodzone_highrisk,
                       data = EP_model_train, 
                       method="rf", 
                       tuneGrid=tunegrid, 
                       trControl=control)
print(rf_gridsearch)
plot(rf_gridsearch)
```

## SVM

```{r set up svm model 1, message=TRUE, warning=TRUE}
svm_reg1 = svm(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                 car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                 road_age + VMT_pop + dist_hydro + dist_major_int + 
                 CLASS + land_use_type + 
                 rb_base + rb_surface +
                 floodzone_highrisk,
               data = EP_model_train, 
               kernel = "polynomial", 
               cost = 10, 
               scale = FALSE)
print(svm_reg1)
```

```{r prediction with svm model 1, message=TRUE, warning=TRUE}
EP_model_svm1 <- 
  EP_model_test %>%
  mutate(svm_predict = predict(svm_reg1, EP_model_test),
         svm_error = svm_predict - PCI_2018,
         svm_absError = abs(svm_predict - PCI_2018),
         svm_APE = abs((svm_predict - PCI_2018)) / PCI_2018) 


MAE <- mean(EP_model_svm1$svm_absError, na.rm = T)
MAPE <- mean(EP_model_svm1$svm_APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

```{r set up svm model 2, message=TRUE, warning=TRUE}
svm_reg2 = svm(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                 car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                 road_age + VMT_pop + dist_hydro + dist_major_int + 
                 CLASS + land_use_type + 
                 rb_base + rb_surface +
                 floodzone_highrisk,
               data = EP_model_train, 
               kernel = "radial", 
               cost = 10, 
               scale = FALSE)
print(svm_reg2)

```

```{r prediction with svm model 2, message=TRUE, warning=TRUE}
EP_model_svm2 <- 
  EP_model_test %>%
  mutate(svm_predict = predict(svm_reg2, EP_model_test),
         svm_error = svm_predict - PCI_2018,
         svm_absError = abs(svm_predict - PCI_2018),
         svm_APE = abs((svm_predict - PCI_2018)) / PCI_2018) 


MAE <- mean(EP_model_svm2$svm_absError, na.rm = T)
MAPE <- mean(EP_model_svm2$svm_APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

```{r set up svm model 3, message=TRUE, warning=TRUE}
svm_reg3 = svm(PCI_2018 ~  crash_len16_18 + potholes_len16_18 + 
                 car_facility_nn3 + entertainment_nn3 + food_drink_nn3 + 
                 road_age + VMT_pop + dist_hydro + dist_major_int + 
                 CLASS + land_use_type + 
                 rb_base + rb_surface +
                 floodzone_highrisk,
               data = EP_model_train, 
               kernel = "sigmoid", 
               cost = 10, 
               scale = FALSE)
print(svm_reg3)

```

```{r prediction with svm model 3, message=TRUE, warning=TRUE}
EP_model_svm3 <- 
  EP_model_test %>%
  mutate(svm_predict = predict(svm_reg3, EP_model_test),
         svm_error = svm_predict - PCI_2018,
         svm_absError = abs(svm_predict - PCI_2018),
         svm_APE = abs((svm_predict - PCI_2018)) / PCI_2018) 


MAE <- mean(EP_model_svm3$svm_absError, na.rm = T)
MAPE <- mean(EP_model_svm3$svm_APE, na.rm = T)
acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)

```

## XGBoost

```{r select features for xgb, message=TRUE, warning=TRUE}
xgb_train <-
  EP_model %>% 
  subset(ind == 1) %>%
  dplyr::select(PCI_2018, crash_len16_18, potholes_len16_18,  
                 car_facility_nn3, entertainment_nn3, food_drink_nn3,  
                 road_age, VMT_pop, dist_hydro, dist_major_int,  
                 CLASS, land_use_type,  
                 #rb_base, rb_surface, 
                 floodzone_highrisk, ind)
xgb_test <-
  EP_model %>% 
  subset(ind == 2) %>%
  dplyr::select(PCI_2018, crash_len16_18, potholes_len16_18,  
                 car_facility_nn3, entertainment_nn3, food_drink_nn3,  
                 road_age, VMT_pop, dist_hydro, dist_major_int,  
                 CLASS, land_use_type,  
                 #rb_base, rb_surface, 
                 floodzone_highrisk, ind)

xgb_all = rbind(xgb_train,xgb_test)
```

```{r one hot encoding, message=TRUE, warning=TRUE}
# one hot encoding
ohe_feats = c('CLASS', 'land_use_type', 'floodzone_highrisk')
dummies <- dummyVars(~ CLASS +  land_use_type + floodzone_highrisk, data = xgb_all)

xgb_ohe <- as.data.frame(predict(dummies, newdata = xgb_all))
xgb_combined <- cbind(xgb_all[,-c(which(colnames(xgb_all) %in% ohe_feats))], xgb_ohe)
```

```{r split train and test set for xgb, message=TRUE, warning=TRUE}
xgb_train <- 
  xgb_combined %>%
  subset(ind == 1)

xgb_test <- 
  xgb_combined %>%
  subset(ind == 2)

x_train <-
  xgb_train %>%
  st_drop_geometry() %>%
  dplyr::select(-ind, -PCI_2018)
y_train <-
  xgb_train %>%
  st_drop_geometry() %>%
  dplyr::select(PCI_2018)

x_test <-
  xgb_test %>%
  st_drop_geometry() %>%
  dplyr::select(-ind, -PCI_2018)
y_test <-
  xgb_test %>%
  st_drop_geometry() %>%
  dplyr::select(PCI_2018)

train_set <- xgb.DMatrix(data = as.matrix(x_train), label = as.matrix(y_train))
test_set <- xgb.DMatrix(data = as.matrix(x_test), label = as.matrix(y_test))
```

```{r train xgb model, message=TRUE, warning=TRUE}
xgb <- xgboost(data = train_set,
               #booster = 'gbtree', 
               eta = 0.15,
               max_depth = 11, 
               nround=1000, 
               subsample = 0.5,
               colsample_bytree = 0.5,
               objective = "reg:squarederror",
)
xgb
```

```{r xgb predict and calculate errors, message=TRUE, warning=TRUE}
xgb_result <- x_test
xgb_result$PCI_2018 <- y_test$PCI_2018
CLASS <- EP_model %>% 
  subset(ind == 2) %>%
  st_drop_geometry() %>%
  dplyr::select(CLASS)
xgb_result$CLASS <- CLASS$CLASS
pave_length <- EP_model %>% 
  subset(ind == 2) %>%
  st_drop_geometry() %>%
  dplyr::select(pave_length)
xgb_result$pave_length <- round(as.numeric(pave_length$pave_length), digits = -1)

land_use_type <- EP_model %>% 
  subset(ind == 2) %>%
  st_drop_geometry() %>%
  dplyr::select(land_use_type)
xgb_result$land_use_type <- land_use_type$land_use_type

floodzone_highrisk <- EP_model %>% 
  subset(ind == 2) %>%
  st_drop_geometry() %>%
  dplyr::select(floodzone_highrisk)
xgb_result$floodzone_highrisk <- floodzone_highrisk$floodzone_highrisk

xgb_result$pred_xgb <- predict(xgb, as.matrix(x_test))
xgb_result$absError <- abs(xgb_result$pred_xgb - xgb_result$PCI_2018)
MAE <- mean(abs(xgb_result$pred_xgb - xgb_result$PCI_2018)) #MAE
MAPE <- mean(abs(xgb_result$pred_xgb - xgb_result$PCI_2018)/xgb_result$PCI_2018) #MAPE

acc <- data.frame(MAE, MAPE)
kable(acc) %>% 
  kable_styling(full_width = F)
```

```{r grid search cv for xgb, message=TRUE, warning=TRUE}
hyper_grid <- expand.grid(max_depth = seq(1, 12, 2), eta = seq(.1, .3, .05))  
xgb_train_rmse <- 0
xgb_test_rmse <- 0

for (j in 1:nrow(hyper_grid)) {
  set.seed(123)
  m_xgb_untuned <- xgb.cv(
    data = train_set,
    nrounds = 1000,
    objective = "reg:squarederror",
    early_stopping_rounds = 3,
    nfold = 5,
    max_depth = hyper_grid$max_depth[j],
    eta = hyper_grid$eta[j]
  )
  
  xgb_train_rmse[j] <- m_xgb_untuned$evaluation_log$train_rmse_mean[m_xgb_untuned$best_iteration]
  xgb_test_rmse[j] <- m_xgb_untuned$evaluation_log$test_rmse_mean[m_xgb_untuned$best_iteration]
  
  cat(j, "\n")
}    


```

# 3. Model Results

## Results for RF

```{r RF prediction result distribution, message=TRUE, warning=TRUE}
EP_model_rf1$pave_length <-
  round(as.numeric(EP_model_rf1$pave_length), digits = -1)
  
EP_model_rf1 %>%
ggplot(aes(CLASS, rf_absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. Road Class",
         y="Absolute Error",
         x="Road Class",
       subtitle = "Random Forest")  + 
  plotTheme()

EP_model_rf1 %>%
ggplot(aes(PCI_2018, rf_absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. PCI",
         y="Absolute Error",
         x="PCI",
       subtitle = "Random Forest")  + 
  plotTheme()

EP_model_rf1 %>%
ggplot(aes(floodzone_highrisk, rf_absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. Flood Zone",
         y="Absolute Error",
         x="Floodzone",
       subtitle = "Random Forest")  + 
  plotTheme()

EP_model_rf1 %>%
ggplot(aes(pave_length, rf_absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  xlim(0, 1500) +
  labs(title = "Absolute Error vs. Pavement Length",
         y="Absolute Error",
         x="Pave length",
       subtitle = "Random Forest")  + 
  plotTheme()

EP_model_rf1 %>%
ggplot(aes(rf_absError, land_use_type)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. Land Use",
         x="Absolute Error",
         y="Land Use Type",
       subtitle = "Random Forest")  + 
  plotTheme()
```

## Results for XGB

```{r XGB prediction result distribution, message=TRUE, warning=TRUE}
xgb_result %>%
ggplot(aes(CLASS, absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. Road Class",
         y="Absolute Error",
         x="Road Class",
       subtitle = "XGBoost")  + 
  plotTheme()

xgb_result %>%
ggplot(aes(PCI_2018, absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. PCI",
         y="Absolute Error",
         x="PCI",
       subtitle = "XGBoost")  + 
  plotTheme()

xgb_result %>%
ggplot(aes(floodzone_highrisk, absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. PCI",
         y="Absolute Error",
         x="Floodzone",
       subtitle = "XGBoost")  + 
  plotTheme()

xgb_result %>%
ggplot(aes(pave_length, absError)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  xlim(0, 1500) +
  labs(title = "Absolute Error vs. Pavement Length",
         y="Absolute Error",
         x="Pave length",
       subtitle = "XGBoost")  + 
  plotTheme()

xgb_result %>%
ggplot(aes(absError, land_use_type)) +
  geom_bar(position = "dodge", stat = "summary", fun = "mean", fill="#5192B0") +  
  labs(title = "Absolute Error vs. Land Use",
         x="Absolute Error",
         y="Land Use Type",
       subtitle = "XGBoost")  + 
  plotTheme()
```

